---
title: "Data processing assignment: Exam examiner"
output: html_notebook
---


## The data

The file `exam_results.xlsx` contains data about some (fictional) students taking modules in a (fictional) Physics department.



## The tasks

0. Load this file into a dataframe. What are the observations and what are the variables?

```{r}
library(tidyverse)
library(readxl)

exam <- read_excel("assets/exam_results.xlsx")
str(exam)
head(exam)
```


1. Tidy the data.

```{r}
# exam <- exam %>%
exam_tidied <- exam %>%
    gather(c("2019", "2020"), key = "Year", value = "Grade", na.rm = TRUE) %>%
    mutate(Year = as.integer(Year)) %>%
    separate(col = "Degree Programme", sep = " with ", into = c("Degree + Major", "Minor")) %>%
    separate(col = "Degree + Major", sep = " ", into = c("Degree", "Major"))

exam_tidied
```


2. Using your tidied data, answer the following questions:

a) For MSci students, what is the mean grade for the Relativity exam?

```{r}
mean_msci_relativity <- exam_tidied %>%
    filter(Degree == "MSci" & Module == "Relativity") %>%
    pull(Grade) %>%
    mean()

mean_msci_relativity
```


b) What proportion of the students are taking Physics as their major subject?

```{r}
is_major_physics <- exam_tidied %>%
    count(Major == "Physics") %>%
    mutate(prop = n / sum(n))

is_major_physics
```



---
