---
title: "Data processing assignment: COVID-19"
output: html_notebook
---


## The data

You are provided with a CSV file, `covid-19.csv` which was downloaded from the EU's open data portal, https://data.europa.eu/euodp/en/home. This file contains data for COVID-19 cases and deaths for each country since the end of 2019.

The dates given in the first column are in the format `dd/mm/yyyy`, e.g. `16/05/2020`.
These can be converted to R's `date` datatype using the `dmy()` function from the tidyverse `lubridate` package:

```{r}
library(tidyverse)
library(lubridate)

dmy("16/05/2020")
```


## The tasks

0. Use the tidyverse functions to load this file into a dataframe and convert the dates from strings to the `date` datatype.

```{r}
covid <- read_csv("assets/covid-19.csv") %>%
    mutate(date = dmy(dateRep))

str(covid)
head(covid)
```

---

Now create figures to show the following:

1. Daily cases for a single country.

```{r}
unique(covid$countriesAndTerritories)
```

```{r}
covid %>%
    filter(countriesAndTerritories == "United_Kingdom") %>%
    ggplot(aes(x = date, y = cases)) +
    geom_line()
```


2. Weekly cases for a single country. *Hint*: use `week()` to convert a date to a number representing the week of the year.

```{r}
covid %>%
    mutate(week = week(date)) %>%
    filter(year == 2020 & countriesAndTerritories == "United_Kingdom") %>%
    group_by(week) %>%
    summarise(weeklyCases = sum(cases)) %>%
    ggplot(aes(x = week, y = weeklyCases)) +
    geom_col()
```


3. Cumulative daily cases for a single country. *Hint*: use `arrange()` and `cumsum()`


```{r}
covid %>%
    filter(countriesAndTerritories == "United_Kingdom") %>%
    arrange(date) %>%
    mutate(totalCases = cumsum(cases)) %>%
    ggplot(aes(x = date, y = totalCases)) +
    geom_line()
```

4. Daily cases for each continent. *Hint*: use `group_by()` and `summarise()`

```{r}
covid %>%
    arrange(date) %>%
    group_by(continentExp, date) %>%
    summarise(dailyContinentCases = sum(cases)) %>%
    ggplot(aes(x = date, y = dailyContinentCases, color = continentExp)) +
    geom_line()
```


5. Apparent overall mortality rate (total deaths / total cases) for each country, grouped by continent.

```{r}
covid %>%
    group_by(continentExp, countriesAndTerritories) %>%
    summarise(totalDeaths = sum(deaths), totalCases = sum(cases)) %>%
    mutate(mortalityRate = totalDeaths / totalCases) %>%
    ggplot(aes(x = continentExp, y = mortalityRate, fill = continentExp)) +
    geom_boxplot()
```

---

6. The UK (Referred to as "United_Kingdom" in this data set) has a strong weekly periodicity in reported COVID-19 deaths.

(a) Make a plot to illustrate this effect as clearly as possible. *Hint*: `lubridate` provides a function `wday()` which returns the day of the week as an integer, starting from 1 = Sunday.

```{r}
name <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
covid %>%
    filter(countriesAndTerritories == "United_Kingdom") %>%
    mutate(weekday = as.character(wday(date))) %>%
    group_by(weekday) %>%
    summarise(weekdayDeaths = sum(deaths)) %>%
    ggplot(aes(x = weekday, y = weekdayDeaths, fill = weekday)) +
    geom_col() +
    guides(fill = "none") +
    scale_x_discrete(labels = name)
```

(b) Show that the number of deaths reported on Mondays and Tuesdays is significantly lower than on other days.

```{r}
name = c("Other", "Monday + Tuesday", "Monday + Tuesday", "Other", "Other", "Other", "Other")
covid %>%
    # filter(year == 2020, month == 10, countriesAndTerritories == "United_Kingdom") %>%
    filter(countriesAndTerritories == "United_Kingdom") %>%
    mutate(week = week(date), weekday = wday(date), dayclass = name[weekday]) %>%
    group_by(week, dayclass) %>%
    summarise(mean_deaths = mean(deaths)) %>%
    spread(key = dayclass, value = mean_deaths) %>%
    filter(`Monday + Tuesday` > 0 | Other > 0)
```

(c) Do any other countries show a similar periodicity?

```{r}
name <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
covid %>%
    filter(countriesAndTerritories == "China") %>%
    mutate(weekday = as.character(wday(date))) %>%
    group_by(weekday) %>%
    summarise(weekdayDeaths = sum(deaths)) %>%
    ggplot(aes(x = weekday, y = weekdayDeaths, fill = weekday)) +
    geom_col() +
    guides(fill = "none") +
    scale_x_discrete(labels = name)
```

(d) How could you explain what is happening, and what are the implications for how we use the data?

Explanations:
- Significant less Mon/Tue cases reported in UK: maybe people don't go out (and infected) on weekends. Test results may have around 2 days delay.
- Significant more Fri cases reported in China: I don't trust their data at all.

Implications:
- Data is part of history and speaks facts. With proper processing, it may reveal general trends that inspire prediction or decision.
---
